A Linear Solution of the Linearity Challenge

### 2025-12-27 paper/related.tex
more related

\citet{Thiemann19} formalises a subset of \GV~\citep{GayVasconcelos10,Wadler14},
a functional language extended with session communication primitives, along with
an interpreter. \citet{CicconePadovani20} have taken inspiration from this work
to formalise a variant of the linear
$\pi$-calculus~\cite{KobayashiPierceTurner99} that supports dependent types, so
as to enable the description of communication protocols whose structure depends
on the content of messages.

@@

@@

    \caption{\label{tab:comparison} A summary of different formalisations of binary session calculi. All sizes are in kb.}
    \begin{minipage}{.9\textwidth}
    \begin{tabular}{@{}r@{~}@{~}c@{~~}c@{~}|@{~}c@{~~}c@{~~}c@{~~}c@{~}|@{~}r@{~~}r@{~~}r@{~~}r@{~~}r@{~}l@{}}

@@

@@

        \footnote{Includes shared channels and the library for algebra of types.}

@@

\Cref{tab:comparison} reports a summary of the aforementioned formalisations and
also our own. The three leftmost columns list the reference papers (sorted by
publication date), the used proof assistant and the formalised calculus.
%
For simplicity, we have grouped the calculi into classes identified by an
acronym. The acronyms \CP~\cite{Wadler14} and
\GV~\cite{GayVasconcelos10,Wadler14} refer to well-known calculi in the
literature on session types. It should be noted that the actual calculus being
formalised usually differs from (typically, is a strict subset of) the one
identified by the acronym, but it retains at least some of its key features.
%
The acronym \SessionPI refers to (variants of) the session-based $\pi$-calculus
presented by \citet{HondaVasconcelosKubo98} while \LinearPI refers to (variants
of) the linear $\pi$-calculus~\cite{KobayashiPierceTurner99}, into which binary
sessions can be encoded using explicit continuation channels as described by
\citet[extended version]{Kobayashi02b} and \citet{DardhaGiachinoSangiorgi17}.
Finally, \SCP is the structural version of \CP introduced by
\citet{SanoKavanaghPientka23} and \Calculus is our calculus.

The four middle columns of \Cref{tab:comparison} report the relevant qualitative
aspects of the formalisations, namely the technique used to enforce the linear
usage of channels, the technique used to encode channels, whether the
formalisation is intrinsically or extrinsically typed, and whether the calculus
makes use of continuation channels. Concerning the use of continuations, the
approaches based directly on the linear $\pi$-calculus (into which sessions can
be encoded) are marked with {\YES} and those based on a calculus with native
sessions are marked with {\NO}. The calculus \SCP is marked with {\MAYBE}
because, while not being strictly based on the linear $\pi$-calculus, it makes
use of explicit continuations for defining the predicates that check the linear
usage of channels. Finally, all the approaches based on \GV are also marked with
\MAYBE. In principle, \emph{none} of these calculi makes use of continuations.
However, \GV is designed in such a way that each operation acting on a channel
$s$ is a function that returns the result of the operation (if any) \emph{along
with the same channel $s$}. In this way, the type of $s$ can be conveniently
``updated'' to take into account the effect of the operation. As observed by
\citet{Padovani17}, this semantics of the communication primitives is virtually
indistinguishable from one making use of explicit continuation channels.

@@

In general, it is difficult to draw firm conclusions on the effectiveness of the
various approaches in addressing the linearity challenge because the
formalisations differ widely for a variety of entangled factors including the
features of calculus, the used proof assistant, the representation of
processes/expressions/channels, the handling of linearity. Also, most
formalisations predate the publication of the linearity
challenge~\cite{CarboneEtAl24}, hence they are not necessarily targeted to
solving that particular problem. However, some of
them~\cite{RouvoetPoulsenKrebbersVisser20,ZalakainDardha21,SanoKavanaghPientka23,ZackonEtAl25}
are explicitly motivated by the complexity arising from the linear usage of
resources/channels and from context splitting in particular.

@@

Finally, it seems that the use of (explicit) continuations affects the
complexity of the formalisation more than anything else. Indeed, the six
smallest formalisations (excluding the deadlock freedom results) -- with an
average size of around 62kb -- are all based on continuations, no matter if they
are explicit (\LinearPI, \SCP, \LCC) or ``virtual'' (\GV), while the remaining
ones are 263kb on average. This observation is in agreement with the
considerations we made earlier in this paper concerning the difference between
\emph{updating} bindings versus \emph{removing and adding} bindings in a typing
context. The use of continuations enables a cleaner management of typing
contexts since channels become true ``use-once'' resources and there is no need
for strong updates.

>>>>

\citet{Thiemann19} formalises a subset of \GV~\citep{GayVasconcelos10}, a
functional language extended with session communication primitives, along with
an interpreter. \citet{CicconePadovani20} have taken inspiration from this work
to formalise a variant of the linear
$\pi$-calculus~\cite{KobayashiPierceTurner99} that supports dependent types, so
as to enable the description of communication protocols whose structure depends
on the content of messages.

@@

    \newcommand\T{\hspace{1.25ex}}

@@

    \caption{\label{tab:comparison} A summary of different formalisations of
    binary session calculi (sizes in kb).}
    \begin{minipage}{.95\textwidth}
    \begin{tabular}{@{}r@{\T}c@{\T}c@{\T}c@{\T}|@{\T}c@{\T}c@{\T}c@{\T}c@{\T}|@{\T}r@{\T}r@{\T}r@{\T}r@{\T}r@{\,}l@{}}

@@

        \textlabel{Cut} &

@@

        \footnote{Includes shared channels and the library for the algebra of types.}

@@

\Cref{tab:comparison} reports a summary of the aforementioned formalisations and
also our own. The four leftmost columns list the reference papers (sorted by
publication date), the used proof assistant, the formalised calculus, and
whether the calculus features a cut in the style of linear logic, as opposed to
separate restriction and parallel composition.
%
For simplicity, we have grouped the calculi into classes identified by an
acronym. The acronyms \CP~\cite{Wadler14} and
\GV~\cite{GayVasconcelos10,Wadler14} refer to well-known calculi in the
literature on session types. The acronym \SessionPI refers to (variants of) the
session-based $\pi$-calculus presented by \citet{HondaVasconcelosKubo98} while
\LinearPI refers to (variants of) the linear
$\pi$-calculus~\cite{KobayashiPierceTurner99}. Finally, \SCP is the structural
version of \CP introduced by \citet{SanoKavanaghPientka23} and \Calculus is our
calculus.
%
It should be noted that the actual calculus being formalised usually differs
from (typically, is a strict subset of) the one identified by the acronym and
that the same acronym may identify different calculi. For example, the first
version of \GV described by \citet{GayVasconcelos10} is not based on linear
logic, whereas later versions~\cite{Wadler14} used \eg by
\citet{JacobsBalzerKrebbers22} are.
%
In any case, the acronym is meant to give an idea of the overall flavour of the
calculus.

The four middle columns of \Cref{tab:comparison} report the relevant qualitative
aspects of the formalisations, namely the technique for the management of typing
contexts, the representation of channels, whether the formalisation is
intrinsically or extrinsically typed, and whether the calculus makes use of
continuation channels. Concerning the use of continuations, the approaches based
directly on the linear $\pi$-calculus (into which sessions can be encoded) are
marked with {\YES} and those based on a calculus with native sessions are marked
with {\NO}. The calculus \SCP is marked with {\MAYBE} because, while not
directly inspired to the linear $\pi$-calculus, it makes use of explicit
continuations for defining the predicates that check the linear usage of
channels. Finally, all the approaches based on \GV are also marked with \MAYBE.
In principle, \emph{none} of these calculi makes use of continuations. However,
\GV is designed in such a way that each operation acting on a channel $s$ is a
function that returns the result of the operation (if any) \emph{along with the
same channel $s$}. In this way, the type of $s$ can be conveniently ``updated''
to take into account the effect of the operation. As observed by
\citet{Padovani17}, this semantics of the communication primitives is virtually
indistinguishable from one making use of explicit continuation channels.

@@

In general, it is difficult to draw firm conclusions on the effectiveness of the
various approaches in addressing the linearity challenge because the
formalisations differ widely for a variety of entangled factors. Also, most
formalisations predate the publication of the linearity
challenge~\cite{CarboneEtAl24}, hence they are not necessarily targeted to
solving that particular problem. However, some of
them~\cite{RouvoetPoulsenKrebbersVisser20,ZalakainDardha21,SanoKavanaghPientka23,ZackonEtAl25}
are explicitly motivated by the complexity arising from the linear usage of
resources/channels and from context splitting in particular.

@@

Using the cut in the style of linear logic instead of separate restriction and
parallel composition simplifies the representation of channels (or session
endpoints). All the formalisations of calculi that adopt the cut tend to be
small (if we exclude the libraries), but this is not a general rule.

Finally, it seems that the use of (explicit) continuations affects the
complexity of the formalisation more than anything else. Indeed, the six
smallest formalisations (excluding the deadlock freedom results) -- with an
average size of around 62kb -- are all based on continuations, no matter if they
are explicit (\LinearPI, \SCP, \LCC) or ``virtual'' (\GV), while the remaining
ones are 263kb on average. This observation is in agreement with the
considerations we made earlier in this paper concerning the difference between
\emph{updating} bindings versus \emph{removing and adding} bindings in a typing
context. The use of continuations enables a cleaner management of typing
contexts since linear channels become true ``use-once'' resources and there is
no need for strong updates of their type.
